---
title: "Exercise 2"
author: "Kyle M. Lang"
date: "Fundamental Techniques in Data Science with R"
params:
  answers: true
output: 
   bookdown::html_document2:
    toc: true
    toc_depth: 1
    toc_float: true
    number_sections: true
---


<style type="text/css">

body{ /* Normal  */
      font-size: 12px;
  }
td {  /* Table  */
  font-size: 12px;
}
h1.title {
  font-size: 18px;
  color: DarkBlue;
}
h1 { /* Header 1 */
  font-size: 18px;
}
h2 { /* Header 2 */
    font-size: 18px;
}
h3 { /* Header 3 */
  font-size: 18px;
}
code.r{ /* Code block */
    font-size: 12px;
}
pre { /* Code block - determines code spacing between lines */
    font-size: 14px;
}
</style>


```{r, echo = FALSE}
knitr::opts_chunk$set(include = params$answers)
```

---

In this practical, you will play around with the different types of R objects. 

---

# Creating Data Objects

---

## 

**Create two vectors**

1. One named `vec1` containing the integers 1 through 6
1. One named `vec2` with letters A through F

```{r}
vec1 <- c(1, 2, 3, 4, 5, 6)
vec2 <- c("A", "B", "C", "D", "E", "F")
```

```{asis, echo = params$answers}
We can use concatenation function, `c()`, to create a vector. A vector is just a 
series of numbers, letters, or other data elements. 
```

---

##

**Create two matrices**

1. One from `vec1`
1. One from `vec2`

Define both matrices to have 3 rows and 2 columns.

```{r}
mat1 <- matrix(vec1, nrow = 3, ncol = 2)
mat2 <- matrix(vec2, nrow = 3, ncol = 2)
```

```{asis, echo = params$answers}
We can use the `matrix()` function to create a matrix. When we create a matrix, 
we need to specify the dimensions (in this case 3 $\times$ 2).
```

---

## 

**Inspect `vec1`, `vec2`, `mat1`, and `mat2`**

Are they all numeric?

```{r}
vec1
vec2
mat1
mat2
```

```{asis, echo = params$answers}
No. `vec1` and `mat1` contain numbers, but `vec2` and `mat2` contain characters. 
```

---

## {#matrixQ}

**Make a 6 $\times$ 2 matrix from both `vec1` and `vec2`**

1. Use `vec1` to populate the first column
1. Use `vec2` to populate the second column

Inspect this matrix.

```{r}
mat3 <- matrix(c(vec1, vec2), 6, 2)
mat3
```

```{asis, echo = params$answers}
OR
```

```{r}
mat3b <- cbind(vec1, vec2)
mat3b
```

```{asis, echo = params$answers}
All elements in a matrix must have the same type. So, if one or more elements in 
the matrix are characters, all other elements are converted to characters.

Notice that the second approach (using the column bind function, `cbind()`) 
returns a matrix with the column names set to the names of the bound objects. 

To create a rectangular dataset that contains different types of data (e.g., 
both numeric and character variables), we can create a data frame. When working 
with real datasets (which often contain diverse data types), data frames are 
usually preferred over matrices. We only really use matrices when doing pure 
numerical calculation, such as in matrix algebra. 
```

---

## {#dfQ1}

**Make a two-column data frame called `dat3`**

1. Use function `data.frame()`.
1. Set `vec1` and `vec2` as the columns. 
1. Name the columns `v1` and `v2`, respectively.

```{r}
dat3 <- data.frame(v1 = vec1, v2 = vec2)
dat3
```

---

# Data Manipulation

---

## {#dfQ2}

**Make a 2-column data frame called `dat3b`**

1. Use the `as.data.frame()` function with the matrix created in Question 
\@ref(matrixQ) as input.
1. Set the argument `stringsAsFactors = TRUE`.
1. Name the columns `v1` and `v2`, respectively. 

Are the types of the `v1` and `v2` columns the same as the types of `vec1` and 
`vec2`, respectively?

```{r}
dat3b <- as.data.frame(mat3, stringsAsFactors = TRUE)
(colnames(dat3b) <- c("v1", "v2"))
```

```{asis, echo = params$answers}
This is a tricky situation. At face value, everything may seem fine, but, we 
need to be careful. The above code may not be doing what you expected. 
Because every element of `mat3` is a character, we have lost the numeric nature 
of `vec1`.
```

---

##

**Check if the first columns in the data frames from Questions \@ref(dfQ1) and 
\@ref(dfQ2) are numeric** 

If these columns are not numeric, determine their type.

```{r}
is.numeric(dat3[, 1])
is.numeric(dat3b[, 1])
```

```{asis, echo = params$answers}
The first column in `dat3b` (from Question \@ref(dfQ2)) is not numeric. As a 
matter of fact, it is not a character vector, either.
```

```{r}
is.character(dat3b[, 1])
```

```{asis, echo = params$answers}
Rather tricky; the function `as.data.frame()` has converted the first variable 
to a factor. 
```

```{r}
is.factor(dat3b[, 1])
```

```{asis, echo = params$answers}
Factors are used to represent categorical variables. Character vectors are not 
converted to factors by default, but the argument `stringsAsFactors = TRUE` 
forces the conversion. Hopefully, you can now see that there is a difference 
between character vectors and factors. Now, you also know how to force the 
conversion from character vectors to factors when creating a data frame. 
```

---

##

**Select the following elements from the data frame you created in Question 
\@ref(dfQ1)**

1. The third row
1. The second column 
1. The intersection of the above

```{r}
dat3[3, ] # 3rd row
dat3[, 2] # 2nd column
dat3$v2   # Also 2nd column
dat3[3, 2] # Intersection
```

```{asis, echo = params$answers}
The matrix-style subsetting operator, `[ , ]`, is very useful in R. For matrices 
and data frames, the first number (before the comma) indexes rows and the 
second number (after the comma) indexes columns. Vectors don't have dimensions, 
so we only need to provide one number for subsetting. For example, `vec1[3]` 
would yield `3`. Try it. 

In data frames, columns can also be extracted with the `$` sign, but only if a 
name has been assigned to the column. If you do not name the columns of your 
data frame, R will assign default column names. 
```

---

##

**Inspect the structure of the data frame that you created in Question 
\@ref(dfQ1).**

The structure function, `str()`, allows us to inspect the structure of an R 
object. Try using it here. 

```{r}
str(dat3)
```

```{asis, echo = params$answers}
Inspecting the structure of your data is vital. If, when we start analyzing our 
data, we assume the wrong measurement level for some of our variables, we may 
run into serious problems. One frequent problem occurs when categorical 
variables are represented on the data as numeric vectors and not factors. 

Notice how the information returned by `str()` mirrors the information in 
RStudio's environment pane. The environment pane is simply reported the ouptut 
from the `str()` function in a beautified format.
```

---

##

**Let's pretend the first variable (`v1`) in the data frame you created in 
Question \@ref(dfQ1) is not coded correctly, and it actually represents grouping 
information about cities.**

Convert the `v1` variable into a factor with the levels:

- Utrecht
- New York, 
- London
- Singapore
- Rome
- Cape Town 

```{r}
dat3$v1 <- factor(dat3$v1, 
                  labels = c("Utrecht", 
                             "New York", 
                             "London", 
                             "Singapore", 
                             "Rome", 
                             "Cape Town")
                  )
dat3
```

---

# Working with Real Data

---

##

**Load the workspace `boys.RData`** 

```{asis, echo = params$answers}
There are a few ways to go about load workspaces that are available on the 
internet. We'll use the *boys.RData* workspace to demonstrate.

1. You can download the *boys.RData* file from the course page and use the 
`load()` function to load it into your environment. 
   - Note that the following code assumes you have saved *boys.RData* in a 
   subdirectory of your project folder called "data".  
```

```{r}
load("data/boys.RData")
```

```{asis, echo = params$answers}
2. You can double-click the `boys.RData` file in your operating system's file 
browser
   - If the file opens in R and not RStudio, you will need to right-click and 
   select the menu option: "Open with" > "RStudio". 

3. You can import workspaces directly from the internet by creating and loading 
a connection.
```

```{r eval = FALSE}
con <- url("https://www.kylemlang.com/fundamentals/data/boys.RData")
load(con)
```

```{asis, echo = params$answers}
In the above code, we store the connection in the object `con`, and then we load 
the connection with `load(con)`.

After executing any of the above options, the `boys` object will be added to 
your Global Environment. You can then use the `boys` data in your analyses.
```

---

Most R packages ship with datasets included (these datasets are most often used 
for examples to demonstrate the functionality of the package). Since you have 
not yet learned how to load packages, you get the `boys` data (which comes from 
the **mice** package) as a stand-alone workspace. 

## 

**View the `boys` dataset two ways**

1. By executing `boys` in the console
1. By using the `View()` function

```{r, eval = FALSE}
boys
View(boys)
```

```{asis, echo = params$answers}
The output is not displayed here as it is simply too large. 

Using `View()` is preferred for inspecting all but the smallest datasets. 
`View()` shows the contents of the dataset in a spreadsheet-like window. 
`View()` is only or viewing the data; you can not edit the dataset's contents 
through the window generated by `View()`.. 
```

---

## 

**Find the dimensions of the `boys` dataset**

```{r}
dim(boys)
```

```{asis, echo = params$answers}
There are `r nrow(boys)` cases on `r ncol(boys)` variables.
```

##

**Inspect the first 6 cases and the final 6 cases in the dataset**

```{r}
boys[1:6, ]
boys[743:748, ]
```

```{asis, echo = params$answers}
OR
```

```{r}
head(boys)
tail(boys)
```

```{asis, echo = params$answers}
The functions `head()` and `tail()` are very useful. For example, by comparing 
the output from these functions, we can infer that the data are very likely 
sorted on `age`.
```

---

##

**Check if the `boys` data are sorted on `age`**

```{asis, echo = params$answers}
To verify if the data are sorted, we can use the `is.unsorted()` function to 
test the inverse of that statement. 

Remember that we can always search the help for functions. For example, we could 
have searched using `?sort` or `??sorted` and quickly found the function 
`is.unsorted()`. 
```

```{r}
is.unsorted(boys$age)
```

```{asis, echo = params$answers}
Since this function returns `FALSE`, we know that `boys$age` is sorted. To 
directly test if `boys$age` is sorted, we could have used:
```

```{r}
!is.unsorted(boys$age)
```

```{asis, echo = params$answers}
This expression tests if `boys$age` is NOT unsorted, so the return value of 
`TRUE` tells us that `boys$age` is sorted. 
```

---

##

**Inspect the `boys` dataset with `str()`**

```{r}
str(boys)
```

---

##

**Use one or more functions to generate numeric summaries of each variable's 
distribution**

At least show the minimum, the maximum, the mean, and the median for all of the 
variables. 

```{r}
summary(boys)
```

---

##

**Give the standard deviations for `age` and `bmi`** 

Tip: Use the help (?) and help search (??) functionality in R, if you get stuck. 

```{r}
sd(boys$age) # SD for age
sd(boys$bmi, na.rm = TRUE) # SD for bmi
```

```{asis, echo = params$answers}
Note that `bmi` contains 21 missing values (you can see these missing values in 
the summary information). Therefore, we need to use `na.rm = TRUE` to calculate 
the standard deviation from the observed cases only.
```

---

# Logical Subsetting

---

##

**Create a new data frame containing only the boys that are 20 years old or older**

How many boys are at least 20 years old?

```{r}
boys2 <- boys[boys$age >= 20, ]
nrow(boys2)
```

```{asis, echo = params$answers}
OR
```

```{r}
boys2 <- subset(boys, age >= 20)
nrow(boys2)
```

```{asis, echo = params$answers}
Logical vectors can be very powerful tools in R. For example, in the first 
solution, we selected the boys that are at least 20 by indexing the rows of the 
data frame with an appropriately defined logical vector.
```

---

##

**Select all boys that are older than 19 but younger than 19.5**

How many boys are between the ages of 19 and 19.5?

```{r}
boys3 <- boys[boys$age > 19 & boys$age < 19.5, ]
nrow(boys3)
```

```{asis, echo = params$answers}
OR
```
 
```{r}
boys3.2 <- subset(boys, age > 19 & age < 19.5)
nrow(boys3.2)
```

---

##

**Compute the mean age of boys younger than 15 years old that do not live in 
region `north`**

```{r}
mean(boys$age[boys$age < 15 & boys$reg != "north" ], na.rm = TRUE)
```

```{asis, echo = params$answers}
OR
```

```{r}
mean(subset(boys, age < 15 & reg != "north")$age, na.rm = TRUE)
```

--- 

In this exercise, you have learned some basic R usage. The approaches we used 
for this exercise offer tremendous flexibility but may also be inefficient in 
complex analyses or data manipulation. Doing advanced operations in basic R can 
require lots of code. In the next exercise, we will start using packages that 
allow us to do more complicated operations with fewer lines of code. 

As you start using R in your own research, you will quickly find yourself in 
need of packages that are not part of the default R installation. The beauty of 
R is that its functionality is community-driven. Anyone can add packages to 
CRAN, and other people can use and improve these packages. There's a good chance 
that a function and/or package has been already developed for the analysis or 
operation you need. If not, maybe you're interested in filling the gap by 
submitting your own package?

---

End of Exercise 2