---
title: "Exercises in Markdown"
author: "Kyle M. Lang"
date: "Fundamental Techniques in Data Science with R"
output: html_document
---

---

1. **Create an object a with value 1**
```{r}
a <- 1
```

--- 

2. **Verify that 1 is stored in a**
```{r}
a
```

---

3. **Square a**

```{r}
a^2
```

---

4. **Assign a + a to the object b, and check if b is equal to a + a.**

```{r}
b <- a + a
b == a + a
```

---

5. 
a. **Square b**
a. **Multiply the answer by a over b**
a. **Assign the result to c**
a. **Take the square root of c to the power b**
a. **Multiply the answer by a over (b to the power 6)**


```{r}
c <- b^2 * (a / b)
sqrt(c^b) * (a / b^6)
```

---

6. **Round the answer from the previous question to 3 decimal places.**

```{r}
round(sqrt(c^b) * (a / b^6), 3)
```

OR

```{r}
d <- sqrt(c^b) * (a / b^6)
round(d, 3)
```

Do you notice that the first solution is shorter, but the second is much more
insightful? If you share your code, or plan to use it later on, clarity very
important. A little effort now, saves you hours later.

---

Now you know how to use R as a calculator and rmarkdown for typesetting.
